---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 2: Group 14"
author: "Helene Behrens, Ellisiv Steen and Johanne Skogvang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document
  #pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
```

```{r,eval=FALSE,echo=TRUE}
install.packages("knitr") #probably already installed
install.packages("rmarkdown") #probably already installed
install.packages("bestglm")# for subset selection with categorical variables
install.packages("glmnet")# for lasso
install.packages("tree") #tree
install.packages("randomForest") #for random forest
install.packages("ElemStatLearn") #dataset in Problem 2
BiocManager::install(c("pheatmap")) #heatmap in Problem 2
```

```{r,eval=FALSE}
install.packages("ggplot2")
install.packages("GGally") # for ggpairs
install.packages("caret") #for confusion matrices
install.packages("pROC") #for ROC curves
install.packages("e1071") # for support vector machines
install.packages("nnet") # for feed forward neural networks
install.packages("MASS")
```

# Problem 1: Regression [6 points]

```{r}
all=dget("https://www.math.ntnu.no/emner/TMA4268/2019v/data/diamond.dd")
dtrain=all$dtrain
dtest=all$dtest
```

**Q1**: Would you choose `price` or `logprice` as response variable? Justify your choice. Next, plot your choice of response pairwise with `carat`, `logcarat`, `color`, `clarity` and `cut`. Comment.
```{r plott}


hist(dtrain$price)
hist(dtrain$logprice)


fit_price = lm(price ~ .-logprice, data=dtrain)
rresprice = rstudent(fit_price)
```

Looking at the histograms of price frequency for the `price` and `logprice` response variable, we see that while the `price` variable have much higher frequenies at lower values, the `logprice` variable is more evenly and symertically distributed for different values. We then want to use `logprice` as our response variable to more easily observe nuances in the data set. 
We can observe the same by looking at Boxcox and residual plots for `price`and `logprice`. 


```{r plott2}
library(ggplot2)
# residuls vs fitted
ggplot(fit_price, aes(.fitted, .stdresid)) + geom_point(pch = 21) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, col = "red", size = 0.5, method = "loess") +
  labs(x = "Fitted values", y = "Standardized residuals",
       title = "Fitted values vs. standardized residuals for price",
       subtitle = deparse(fit_price$call))

library(MASS)
boxcox(fit_price, plotit=TRUE)

fit_logprice = lm(logprice ~.-price, data=dtrain)
library(ggplot2)
# residuls vs fitted
ggplot(fit_logprice, aes(.fitted, .stdresid)) + geom_point(pch = 21) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, col = "red", size = 0.5, method = "loess") +
  labs(x = "Fitted values", y = "Standardized residuals",
       title = "Fitted values vs. standardized residuals for logprice",
       subtitle = deparse(fit_logprice$call))
library(MASS)
boxcox(fit_logprice, plotit=TRUE)


```
We see that in the BoxCox plots, the lambda value at maximum log-likelihood is closer to one for the logprice than for price. When lambda is close to one it means that one should use the model that is plotted. Also, we want the residuals to be independent of each other, meaning they are randomly spread and that we do not see a pattern in the residual plot. We also we want their variance to be constant. Therefore we choose to use the logprice as responce variable. 

We now plot `logprice` with `carat`, `logcarat`, `color`, `clarity` and `cut`.

```{r }
plot(dtrain$carat, dtrain$logprice, main = 'Carat vs logprice')
plot(dtrain$logcarat, dtrain$logprice, main = 'Logcarat vs logprice')
plot(dtrain$clarity, dtrain$logprice, main = 'Clarity vs logprice')
plot(dtrain$cut, dtrain$logprice, main = 'Cut vs logprice')
```

From both the plot with `carat` and `logcarat` we observe a clear indication that higher carat leads to a higher price, which is expected. The relation between `logcarat` and `logprice` seems to be linear, which indicates that the price grows linearly with the carat. 

The relation between `logprice`and `clarity` is less apparent. Looking at the medians and the mid 50 percentiles, is seems as though the price decreses with better clarity, which is counterintuitive. We do however, observe that many outliers in the most extreme observations, and that the whiskers for most of the clarities cover mostly the entire price range. This indicates that the clarity has little influence on the diamond price. 

Some of the same is observable in the plot between `cut` and `logprice`, where the whiskersof the boxplots covers most of the price range for all cuts. 

Use the local regression model $\texttt{logprice} = \beta_0 + \beta_1 \texttt{carat}+ \beta_2 \texttt{carat}^2$ weighted by the tricube kernel $K_{i0}$. 

**Q2:** What is the predicted price of a diamond weighting 1 carat. Use the closest 20% of the observations.  
```{r message=FALSE, warning=FALSE}
#fit = loess(dtrain$logprice ~ dtrain$carat, span=.2)
fit = loess(logprice ~ carat, data=dtrain, span=.2 )
10^predict(fit, 1)

```

**Q3:** What choice of $\beta_1$, $\beta_2$ and $K_{i0}$ would result in KNN-regression?  
In local regression we have $\hat{f} = \beta_0+\beta_1 x_i+\beta_2x_i^2 $, while for KNN-regression this $\hat{f}$ is konstant equal to the mean of the observations in the neighbourhood. Hence, for this to be the KNN-regression we have $ \beta_1 = \beta_2 = 0$. Also, we want $K_{i0}$ equal to one for the observations in the neighbourhood, and zero for the ones outside. 

**Q4:** Describe how you can perform model selection in regression with AIC as criterion.  
When we perform model selection in regression with AIC as the criterion, we perform a best subset selcetion, and choose the set of covariates that produce the lowest AIC value. 
The full procedure when one has a model depending on $k$ covariates is as follows:
1) For every value $i$ from 1 to $k$, all $k \choose j$ models that contain $j$ covariates are fitted. 
2) The model $M_j$ with the lowest SSE value is selected. 
3) From all $M_j$, $j  = 0, ... , k$, where $M_0$ is the model with only the intercept, the model with the lowest AIC is selected. The AIC is defined as

$$\text{AIC} = \frac{\text{SSE}/n}{\hat{\sigma^2}} + 2\frac{p}{n}$$
where $n$ is the number of observations and $p$ is the number of selected covariates in the given model. 

**Q5:** What are the main differences between using AIC for model selection and using cross-validation (with mean squared test error MSE)?  

Cross-validation is in general more computationally expensive than model selection with AIC, however, this is rearly considered a problem with today's computers. Cross-validation also measures the test error directly, instead of adjusting the training error for model size, as the AIC does. Finally, cross-validation makes fewer assumptions about the underlying model. Cross-validation does not directly punish models with many covariates; if two models with different complexity produces the same test error, cross-validation will do nothing to choose the less complex model. 

**Q6:** See the code below for performing model selection with `bestglm()` based on AIC. What kind of contrast is used to represent `cut`, `color` and `clarity`? Write down the final best model and explain what you can interpret from the model.  

**Q7:** Calculate and report the MSE of the test set using the best model (on the scale of `logprice`).  

```{r,eval=FALSE}
library(bestglm)
ds=as.data.frame(within(dtrain,{
  y=logprice    # setting reponse
  logprice=NULL # not include as covariate
  price=NULL    # not include as covariate
  carat=NULL    # not include as covariate
  }))
fit=bestglm(Xy=ds, IC="AIC")$BestModel
summary(fit)

predict(fit,1)

#NB! dette svarer ikke på Q7!
res_fit <- fit$residuals
abs_res <- abs(res_fit)
n <- length(res_fit)
sum_res <- sum(abs_res**2)
sum_res
mse <- sum_res / n
mse

```

**Q8:** Build a model matrix for the covariates `~logcarat+cut+clarity+color+depth+table+xx+yy+zz-1`. What is the dimension of this matrix?  

**Q9:** Fit a lasso regression to the diamond data with `logprice` as the response and the model matrix given in Q8. How did you find the value to be used for the regularization parameter?  

**Q10:** Calculate and report the MSE of the test set (on the scale of `logprice`).  

**Q11:** A regression tree to model is built using a _greedy_ approach. What does that mean? Explain the strategy used for constructing a regression tree.  
**Q12:** Is a regression tree a suitable method to handle both numerical and categorical covariates? Elaborate.  
**Q13:** Fit a (full) regression tree to the diamond data with `logprice` as the response (and the same covariates as for c and d), and plot the result. Comment briefly on you findings.  
**Q14:** Calculate and report the MSE of the test set (on the scale of `logprice`).  

**Q15:** Explain the motivation behind bagging, and how bagging differs from random forest? What is the role of bootstrapping?  
**Q16:** What are the parameter(s) to be set in random forest, and what are the rules for setting these?  
**Q17:** Boosting is a popular method. What is the main difference between random forest and boosting?  
**Q18:** Fit a random forest to the diamond data with `logprice` as the response (and the same covariates as before). Comment on your choice of parameter (as decribed in Q16).   
**Q19:** Make a variable importance plot and comment on the plot. Calculate and report the MSE of the test set (on the scale of `logprice`).  

**Q20:** Finally, compare the results from c (subset selection), d (lasso), e (tree) and f (random forest): Which method has given the best performance on the test set and which method has given you the best insight into the relationship between the price and the covariates?


# Problem 2: Unsupervised learning [3 points]

**Q21:** The principal component score is a measurement of how important a given set of observations is for a principal component. The principal components are the eigenvectors of the covariance matrix ${\hat {\bf R}}$, and thus a principal component score for a set of observations is the scaled observations multiplied with a principal component.
$$\text{score}_i = \sum_{j=1}^{p}x_{ij}\cdot(PCi_j)$$
```{r}
library(ElemStatLearn)
X=t(nci) #n times p matrix
ngroups=length(table(rownames(X)))
cols=rainbow(ngroups)
cols[c(4,5,7,8,14)] = "black"
pch.vec = rep(4,14)
pch.vec[c(4,5,7,8,14)] = 15:19

colsvsnames=cbind(cols,sort(unique(rownames(X))))
colsamples=cols[match(rownames(X),colsvsnames[,2])] 
pchvsnames=cbind(pch.vec,sort(unique(rownames(X))))
pchsamples=pch.vec[match(rownames(X),pchvsnames[,2])]

Z=scale(X)

pca=prcomp(Z)
plot(pca$x[,1],pca$x[,2],xlab="PC1",ylab="PC2",pch=pchsamples,col=colsamples)
legend("bottomright",legend = colsvsnames[,2],cex=0.55,col=cols,pch = pch.vec)

plot(1:dim(X)[2],pca$rotation[,1],type="l",xlab="genes",ylab="weight",main="First eigenvector")

```


**Q22:** The plot "First eigenvector" shows the weights of the first eigenvector of $\hat{\bf{R}}$ which is also the first principal component. The reason why there is only $n=64$ principal component is that there can not be more principal components than the rank of $\bf{X}$. This is because all the principal components are linearly independent, and thus the number of principal components (which is the eigenvectors of $\hat{\bf{R}}$) is bounded by the rank of $X$ which is maximally $n - 1=63$ (reduces to $n-1$ because of the centering). As seen from **Q21**, the score is directly related to a given principal component, so the number of component scores is always the same as the number of principle components.

**Q23:** The proportion of variance explained by a single principle component is calculated as 
$$\text{PVE}=\frac{\text{variance explained by }m\text{th component}}{\text{total variance}} = \frac{\frac{1}{n}\sum_{i=1}^{n}(\sum_{j = 1}^{p} \phi_{jm}z_{ij})^2}{\frac{1}{n}\sum_{j=1}^{p}\sum_{i=1}^{n}z_{ij}}.$$
From this formula it is also easy to see that the total variance `sum(pca$sdev^2=p)`. The total variance defined above is the same as the sum of the variances of the columns of $\bf{Z}$ which is $p$ because they are standardized and thus every column has variance $1^2=1$. An easier way of computing the proportion of variance is to use the eigenvalues corresponding to the principle components. It can be shown that $\text{Var}(\phi_j) = \lambda_j$ and thus $$\text{PVE}=\frac{\lambda_m}{\sum_{j=1}^{p}\lambda_j}$$

```{r proportion of variance}
pve=100*pca$sdev^2/sum(pca$sdev^2)
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal component", col="brown3")

totvar = sum(pca$sdev^2)
pve32=sum(pca$sdev[1:32]^2)/totvar

pve31 = sum(pca$sdev[1:31]^2)/totvar

cat("The first 31 principal components explains", pve31 * 100, "% of the variance and 32 principal components explains", pve32 * 100, "% of the variance. Thus we need 32 principal components to explain 80% of the variance in the data.")
```

**Q24**: From the plot above, plotting PC1 agianst PC2, the most obvious clustering is the blue clustering, where the genes are from `MELANOMA` cancer tumors. The green `LEUKEMIA` genes are also easy to seperate from the others, and so are genes from `COLON` cancer tumors. We observe that the genes from other cancer types are mostly located in the left upmost corner and is for the most grouped together. The exeption is the genes from the breast cancer tumor, which are not clustered together at all and is spread around the whole plot. 
  It is known that the `K562` samples are leukemia cells and we can see from the plot that they are clustered together with the other leukemia samples. The breast cancer samples `MCF7` is harder to say something about because the other breast cancer samples are also spread widely around. The unknown sample is laying in the upper left corner, so it could belong to any of those group really. The most likely cancer type is although `NSCLC` or `OVARIAN` because the closest points are of those types.

We plot a couple of other projections of the data onto the principal components to see if we find other interesting relationships. It is mostly interesting to look at the fist principal comonents, because it is in those that contains the most variations. It could have been interesting to find plots where the breast cancer cells are clustered together, but this was very hard to do. Below two other pairs of principal components are plotted together. In the first plot we see PC1 agianst PC3 and here we also see that green and yellow are clustered together and in the second we see the blue points clustered clearly together. After plotting several principal components, we see that it is easy to separate the blue, green and yellow points, but other than that it is hard to find PCs clearly separating other colors.

```{r}
plot(pca$x[,1],pca$x[,3],xlab="PC1",ylab="PC3",pch=pchsamples,col=colsamples)
legend("bottomright",legend = colsvsnames[,2],cex=0.55,col=cols,pch = pch.vec)

plot(pca$x[,2],pca$x[,4],xlab="PC2",ylab="PC4",pch=pchsamples,col=colsamples)
legend("bottomright",legend = colsvsnames[,2],cex=0.55,col=cols,pch = pch.vec)
```


**Q25:**: Explain what it means to use Euclidean distance and average linkage for hierarchical clustering. 

**Q26:**: Perform hierarchical clustering with Euclidean distance and average linkage on the scaled gene expression in `Z`. Observe where our samples labelled as K562, MCF7 and `UNKNOWN` are placed in the dendrogram. Which conclusions can you draw from this?

**Q27:**: Study the R-code and plot below. Here we have performed hierarchical clustering based on thefirst 64 principal component instead of the gene expression data in `Z`. What is the difference between using all the gene expression data and using the first 64 principal components in the clustering? We have plotted the dendrogram together with a heatmap of the data. Explain what is shown in the heatmap. What is given on the horizontal axis, vertical axis, value in the pixel grid?

```{r}
library(pheatmap)
npcs=64 
pheatmap(pca$x[,1:npcs],scale="none",cluster_col=FALSE,cluster_row=TRUE,clustering_distance_rows = "euclidean",clustering_method="average",fontsize_row=5,fontsize_col=5)
```

# Problem 3: Flying solo with diabetes data [6 points]

```{r}
flying=dget("https://www.math.ntnu.no/emner/TMA4268/2019v/data/flying.dd")
ctrain=flying$ctrain
ctest=flying$ctest
```

**Q28:** Start by getting to know the _training data_, by producing summaries and plots. Write a few sentences about what you observe and include your top 3 informative plots and/or outputs.

**Q29:** Use different methods to analyse the data. In particular use 

* one method from Module 4: Classification
* one method from Module 8: Trees (and forests)
* one method from Module 9: Support vector machines and, finally
* one method from Module 11: Neural networks

For each method you 

* clearly write out the model and model assumptions for the method
* explain how any tuning parameters are chosen or model selection is performed
* report (any) insight into the interpretation of the fitted model
* evaluate the model using the test data, and report misclassifiation rate (cut-off 0.5 on probability) and plot ROC-curves and give the AUC (for method where class probabilities are given).

**Q30:** Conclude with choosing a winning method, and explain why you mean that this is the winning method.
